Percussive Synthesizer Generation with Virtual Listeners

### 2020-10-11 templates_aimusic2020/paper.tex
Updates from Overleaf

Furthermore, we provide a subset of our curated datasets to facilitate future research. We curated 3 datasets of one-shot drums which we use to train and validate drum classifiers. The number of samples per group in these databases is provided in Table ~\ref{table:drumdb}. We provide unrestricted access to \textbf{FreeDB} via Zenodo\footnote{\url{https://zenodo.org/record/3994999}}.
\textbf{RadarDB} cannot be shared directly, we provide the script for its automated creation\footnote{\url{https://github.com/imilas/Synths_Stacks_Search}}. We cannot provide a copy of \textbf{MixedDB}, which is curated from various sources and personal libraries. 
 % \section{Datasets}
% We curated 3 data-sets of one-shot drums which we use to train drum classifiers. \textbf{FreeDB} contains 5 categories: kicks, snares, claps, hats and other (various percussions). \textbf{RadarDB} contains 7 categories: kicks, snares, claps, hats, shakers, toms and rims. \textbf{MixedDB} contains 8 categories: kicks, snares, claps, high toms, low toms, medium toms, closed hats and open hats. See Appendix~\ref{appendix:datasets} for details and download links.   
\section{Related Works}
Numerous deep neural network models have been proposed for the purpose of signal generation in recent years~\cite{nsynth2017,yamamoto2020parallel,oord2017parallel,yee2018automatic,ramires2020neural}. WaveGans and WaveNet have been subject to significant improvements and experiments since their proposal~\cite{nsynth2017,yamamoto2020parallel,oord2017parallel}. Particularly relevant works are the utilization of variational autoencoders (VAE's) for generation of percussive samples~\cite{aouameur2019neural} and generation of percussive sounds by decoding a small set of latent features~\cite{ramires2020neural}. Automatic programming~\footnote{unsupervised generation of source code or programs towards a goal} of virtual synthesizers has long been a topic of interest. Genetic Algorithms (GA's) have been utilized for the generation of new sounds with various sound-engines~\cite{hornermachinetongues,macret2012automatic}. A more recent work by Yee-King et al.~\cite{yee2018automatic} used Long short-Term Memory (LSTM) models and heuristic search to find the exact parameters used to create a group of sounds. The sounds approximated were made by the same virtual synthesizer and not with an external source. Esling et al. used  over 10,000 VST synthesizer presets to learn a parameter space which can be sampled for creation of new audio~\cite{esling2019universal}. Here, we work towards the approximation of percussive sounds with no prior knowledge about the parameter space of a synthesizer.

@@

\label{table:submodule_params}
\end{table}
To create sounds, we build digital synthesizers. We use classical DSP to build our synthesizer, which allows for quick, offline, and parallel generation of audio signals without the usage of GPUs. We made extensive use of Pippi\footnote{https://github.com/luvsound/pippi} and SciPy~\cite{jones2001scipy} libraries. Our virtual synthesizer contains a set of one or more submodules. Each submodule is a self-contained noise making unit and creates signals by taking the steps depicted in figure~\ref{fig:submodule}. Submodules have identical sets of parameters, but widely different outputs can be achieved depending on the values assigned. The set of required parameters for each submodule is highlighted in Table~\ref{table:submodule_params}. The sonic output of the virtual synthesizer is the normalized addition of the output of its submodules. Our implementation of a synthesizer can have any number of submodules. We call the number of submodules in each virtual synthesizer the \textit{stack size}. We call the sets of parameter values that characterize a synthesizer's submodules a \textit{program} (analogous to presets and submodules for a VST).  
 \begin{figure}[tb]
    \begin{center}
    % \textbf{Synthesizer SubModule }
    \makebox[\textwidth]{
    \includegraphics[width=1\linewidth]{templates_aimusic2020/images/synthesizer_block_thin.pdf}}
    \end{center}
    \caption{High level representation of pre-rendering steps for each submodule. Each Synthesizer contains 1 or more these submodules. Synthesizer programs set the number of these submodules and their parameters. The output of a synthesizer is the normalized addition of all its submodule outputs. 
    }
\label{fig:submodule}
\end{figure}
 Rather than directly using neural networks for sound synthesis, we generate programs for this virtual synthesizer. Our decision is based on the following factors:
\begin{enumerate*}[label=(\roman*)]
    \item \textit{Novelty and Creativity}: The goal here is to work with the limitations of any tractable sound source to create its approximations of a given sound category. We seek to create novel sounds via artificial, exploratory creativity. Boden defines this concept as an emergent property of generative work within confined rule sets~\cite{boden2009computer}. An example is the perpetual popularity of 8-bit aesthetics~\cite{collins2007loop}. 
    \item \textit{Interpretability}: Neural networks are often described as black boxes with uninterruptible weights~\cite{basheer2000artificial}. Their highly recursive structure makes modern explanation methods such as saliency maps unreliable~\cite{rudin2019stop}.  
    \item \textit{Speed of Rendering}: Neural network synthesis is costly; Sub 24 khz sample rates are common in most relevant works~\cite{yamamoto2020parallel,oord2017parallel,aouameur2019neural,ramires2020neural}. This is far below CD quality sampling rates~\cite{reiss2016meta}. At our fixed sampling rate of 48 khz, synthesizers with 8 submodules can create and save 1 second sounds to hard-disk with an average rendering time of 50 milliseconds\footnote{Using a single process on a Macbook Air 2012 and Ubuntu 18.04}. 
    \item \textit{Flexibility and Scaling}: Probabilistic audio generation is often done sequentially. State of the art, parallel wave generation with GANs requires a fixed amount for rendering time for each time-step~\cite{yamamoto2020parallel}. With our virtual synthesizer, the added footprint of increasing the length of rendered sounds or higher sampling rates is relatively minuscule.  
\end{enumerate*}

@@

% Please add the following required packages to your document preamble:
\begin{table}[tb]
\caption{Overview of model architectures. Further details in Appendix~\ref{appendix:classifier_definitions}}
\resizebox{\textwidth}{!}{%
    \begin{tabular}{|>{\centering\arraybackslash}p{2cm}|>{\centering\arraybackslash}p{10cm}|>{\centering\arraybackslash}p{2cm}|}
    \hline
    Model & Architecture Layout                                                                             & Features                 \\ \hline
    CNN(TPE)   & CNN(2 channel convolution)~ -\textgreater~LSTM(800 hidden states)~-\textgreater\newline ~Linear Layers (shape 400x5)                                 & Spectrogram          \\ \hline
    FC(TPE)    & Fully connected Linear layers with shape: 400x10x5x10x5                                & Spectrogram          \\ \hline
    E+F(TPE)   & Fully connected network of size 50x10x2x5 & Env. + Freq. \\ \hline
    MEM   & Encoder -\textgreater~Embedded Spectrogram(size 64) -\textgreater~Extra-Trees                                              & Spectrogram          \\ \hline
    \end{tabular}
}
\end{table}

To automatically learn from the extracted features, two groups of virtual ears are implemented: \emph{two phased ears} (TPEs) and \emph{Mixed Ear Models} (MEMs).  For TPEs, we train multiple neural network architectures using different subsets of the FFT features to specialize in \decfirst~or \decsecond~ and combine them to make decisions sequentially. To train TPEs for \decfirst, we use all drums in RadarDB and FreeDB and 6000 examples of virtual synthesizer noise. For \decsecond, we combine the two databases and merge toms into kicks and rims/shakers into \enquote{other}. We trained the TPE models with 80\% of this dataset. Using the remaining 20\% of sounds, we achieved 98\% accuracy in \decfirst~and 82\% accuracy in~\decsecond~with our best models. These accuracy numbers are weak as we did not account for category sizes or cross validate.

@@

MEMs learn from autoencoder embeddings and envelope features and give simultaneous answers to both decisions. We use kicks, snares, claps, and hat sounds from RadarDB and FreeDB as well as 1000 examples of synthesized noise to train MEMs. Using 10-fold cross validation, our best MEM (an ExtraTrees Classifier) achieves the average f1-score of 92\% in 5-way sound classification (4 drum categories + synthesized noise). Detailed measurements and model descriptions for TPEs and MEMs can be found in Appendix~\ref{appendix:classifier_definitions}.

@@

High accuracy in detection of organic drums vs synthetic sounds does not necessarily make an ear more suitable for our generative system. Perfect detection of organic drums vs synthetic noise would create a system which generates nothing.  We cannot know the true size of $\mathcal{H^{+}}$ (see Figure~\ref{fig:ven_data}) without exhaustive manual hearing tests. We conduct blinded hearing tests of two generative systems. For \decfirst, a network which learns from spectrogram data utilizing CNN and LSTM layers is used. For \decsecond, we categorize the drum types with 3 different models: FC (Fully-Connected architecture and spectrograms for learning), CNN (CNN and LSTM layers with spectrograms), and E/F (Fully-Connected architecture with frequency and envelope features). The MEM system uses our best MEM, which simultaneously classifies sounds as drums and categorizes them.

@@

The percentage of sounds labeled as \enquote{bad} by the authors is a measurement of success with regards to \decfirst. We measure success with regards to \decsecond---the reliability of agreement between persons and drum categorization models---via the Fleiss' kappa coefficient \cite{fleiss1971measuring}. The value of 0 or less for this coefficient indicates no agreement beyond random chance, and the values of 1 indicates perfect agreement. We re-measure this coefficient after dropping all \enquote{bad} and \enquote{other} samples.

@@

\section{Dataset Details and Downloads}
\label{appendix:datasets}
\subsection{Download Instructions}
FreeDB, survey data, and randomly generated sounds can be downloaded from: \url{https://zenodo.org/record/3994999}\\\\
RadarDB requires lengthy downloads and processing. The script for its automated creation can be found under the \enquote{getting\_data} directory of the project:\\
\url{https://github.com/imilas/Synths_Stacks_Search}. 
\\\\
Also included on github is our source-code, models, and extra output examples.

\subsection{Details}
Our data sources are a large set of 2 second or shorter drum samples aggregated from personal libraries, free drum kits from the sample-swap project \footnote{https://sampleswap.org/} which we organized into 5 groups, and a large set of drum sounds aggregated from royalty free sources such as music radar \footnote{https://www.musicradar.com/}. We put together 3 databases of drums using these sources. We also created a database of synthetic noise from 1, 3, and 5 stacked virtual synthesizers. Specifications about our datasets can be found in tables~\ref{table:all_db}.

@@

Using the optuna optimization tools \cite{akiba2019optuna}, we conducted 500 search trials. The trial's success is measured in their final loss value, calculated by applying the model to test data-set. Each trial consisted of 20 epochs of training.

>>>>

Furthermore, we provide a subset of our curated datasets to facilitate future research. We curated 3 datasets of one-shot drums which we use to train and validate drum classifiers. The number of samples per group in these databases is provided in Table ~\ref{table:drumdb}. We provide unrestricted access to \textbf{FreeDB} via Zenodo\footnote{\url{https://zenodo.org/record/3994999}}.
\textbf{RadarDB} cannot be shared directly, we provide the script for its automated creation\footnote{\url{https://github.com/imilas/Synths_Stacks_Search}}. We cannot provide a copy of \textbf{MixedDB}, which is curated from various sources and personal libraries. 
 % \section{Datasets}
% We curated 3 data-sets of one-shot drums which we use to train drum classifiers. \textbf{FreeDB} contains 5 categories: kicks, snares, claps, hats and other (various percussions). \textbf{RadarDB} contains 7 categories: kicks, snares, claps, hats, shakers, toms and rims. \textbf{MixedDB} contains 8 categories: kicks, snares, claps, high toms, low toms, medium toms, closed hats and open hats. See Appendix~\ref{appendix:datasets} for details and download links.   
\section{Related Works}
Numerous deep neural network models have been proposed for the purpose of signal generation in recent years~\cite{nsynth2017,yamamoto2020parallel,oord2017parallel,yee2018automatic,ramires2020neural}. WaveGans and WaveNet have been subject to significant improvements and experiments since their proposal~\cite{nsynth2017,yamamoto2020parallel,oord2017parallel}. Particularly relevant works are the utilization of variational autoencoders (VAE's) for generation of percussive samples~\cite{aouameur2019neural} and generation of percussive sounds by decoding a small set of latent features~\cite{ramires2020neural}. Automatic programming~\footnote{unsupervised generation of source code or programs towards a goal} of virtual synthesizers has long been a topic of interest. Genetic Algorithms (GA's) have been utilized for the generation of new sounds with various sound engines~\cite{hornermachinetongues,macret2012automatic}. A more recent work by Yee-King et al.~\cite{yee2018automatic} used Long short-Term Memory (LSTM) models and heuristic search to find the exact parameters used to create a group of sounds. The sounds approximated were made by the same virtual synthesizer and not with an external source. Esling et al. used  over 10,000 VST synthesizer presets to learn a parameter space which can be sampled for creation of new audio~\cite{esling2019universal}. Here, we work towards the approximation of percussive sounds with no prior knowledge about the parameter space of a synthesizer.

@@

\label{table:submodule_params}
\end{table}
To create sounds, we build digital synthesizers. We use classical DSP to build our synthesizer, which allows for quick, offline, and parallel generation of audio signals without the usage of GPUs. We made extensive use of Pippi\footnote{https://github.com/luvsound/pippi} and SciPy~\cite{jones2001scipy} libraries. Our virtual synthesizer contains a set of one or more submodules. Each submodule is a self-contained noise making unit and creates signals by taking the steps depicted in figure~\ref{fig:submodule}. Submodules have identical sets of parameters, but widely different outputs can be achieved depending on the values assigned. The set of required parameters for each submodule is highlighted in Table~\ref{table:submodule_params}. The sonic output of the virtual synthesizer is the normalized addition of the output of its submodules. Our implementation of a synthesizer can have any number of submodules. We call the number of submodules in each virtual synthesizer the \textit{stack size}. We call the sets of parameter values that characterize a synthesizer's submodules a \textit{program}(analogous to presets and submodules for a VST).  
 \begin{figure}[tb]
    \begin{center}
    % \textbf{Synthesizer SubModule }
    \makebox[\textwidth]{
    \includegraphics[width=1\linewidth]{templates_aimusic2020/images/synthesizer_block_thin.pdf}}
    \end{center}
    \caption{High level representation of pre-rendering steps for each submodule. Each Synthesizer contains 1 or more these submodules. Synthesizer programs set the number of these submodules and their parameters. The output of a synthesizer is the normalized addition of all its submodule outputs. 
    }
\label{fig:submodule}
\end{figure}
 Rather than directly using neural networks for sound synthesis, we generate programs for this virtual synthesizer. Our decision is based on the following factors:
\begin{enumerate*}[label=(\roman*)]
    \item \textit{Novelty and Creativity}: The goal here is to work with the limitations of any tractable sound source to create its approximations of a given sound category. We seek to create novel sounds via artificial, exploratory creativity. Boden defines this concept as an emergent property of generative work within confined rule sets~\cite{boden2009computer}. An example is the perpetual popularity of 8-bit aesthetics~\cite{collins2007loop}. 
    \item \textit{Interpretability}: Neural networks are often described as black boxes with uninterruptible weights~\cite{basheer2000artificial}. Their highly recursive structure makes modern explanation methods such as saliency maps unreliable~\cite{rudin2019stop}.  
    \item \textit{Speed of Rendering}: Neural network synthesis is costly; Sub 24 khz sample rates are common in most relevant works~\cite{yamamoto2020parallel,oord2017parallel,aouameur2019neural,ramires2020neural}. This is far below CD quality sampling rates~\cite{reiss2016meta}. At our fixed sampling rate of 48 khz, synthesizers with 8 submodules can create and save 1 second sounds to hard-disk with an average rendering time of 50 milliseconds\footnote{Using a single process on a Macbook Air 2012 and Ubuntu 18.04}. 
    \item \textit{Flexibility and Scaling}: Probabilistic audio generation is often done sequentially. State of the art, parallel wave generation with GANs requires a fixed amount of rendering time for each time-step~\cite{yamamoto2020parallel}. With our virtual synthesizer, the added footprint of increasing the length of rendered sounds or higher sampling rates is relatively minuscule.  
\end{enumerate*}

@@

% Please add the following required packages to your document preamble:
\begin{table}[tb]
\caption{Overview of survey model architectures. Further details in Appendix~\ref{appendix:classifier_definitions}}
\resizebox{\textwidth}{!}{%
    \begin{tabular}{|>{\centering\arraybackslash}p{2cm}|>{\centering\arraybackslash}p{10cm}|>{\centering\arraybackslash}p{2cm}|}
    \hline
    Model & Architecture Layout                                                                             & Features                 \\ \hline
    CNN(TPE)   & CNN(2 channel convolution)~ -\textgreater~LSTM(800 hidden states)~-\textgreater\newline ~Linear Layers (shape 400x5)                                 & Spectrogram          \\ \hline
    FC(TPE)    & Fully connected Linear layers with shape: 400x10x5x10x5                                & Spectrogram          \\ \hline
    E+F(TPE)   & Fully connected network of size 50x10x2x5 & Env. + Freq. \\ \hline
    MEM   & Encoder -\textgreater~Embedded Spectrogram(size 64) -\textgreater~Extra-Trees                                              & Spectrogram          \\ \hline
    \end{tabular}
}
\label{table:brief_models}
\end{table}

To automatically learn from the extracted features, two groups of virtual ears are implemented: \emph{two phased ears} (TPEs) and \emph{Mixed Ear Models} (MEMs).  For TPEs, we train multiple neural network architectures using different subsets of the FFT features to specialize in \decfirst~or \decsecond~and combine them to make decisions sequentially. To train TPEs for \decfirst, we use all drums in RadarDB and FreeDB and 6000 examples of virtual synthesizer noise. For \decsecond, we combine the two databases and merge toms into kicks and rims/shakers into \enquote{other}. We trained the TPE models with 80\% of this dataset. Using the remaining 20\% of sounds, we achieved 98\% accuracy in \decfirst~and 82\% accuracy in~\decsecond~with our best models. These accuracy numbers are weak as we did not account for category sizes or cross validate.

@@

MEMs learn from autoencoder embeddings and envelope features and give simultaneous answers to both decisions. We use kicks, snares, claps, and hat sounds from RadarDB and FreeDB as well as 1000 examples of synthesized noise to train MEMs. Using 10-fold cross validation, our best MEM (an ExtraTrees Classifier) achieves the average f1-score of 92\% in 5-way sound classification (4 drum categories + synthesized noise). Detailed measurements and model descriptions for TPEs and MEMs can be found in table~\ref{table:brief_models} and Appendix~\ref{appendix:classifier_definitions}.

@@

High accuracy in detection of organic drums vs synthetic sounds does not necessarily make an ear more suitable for our generative system. Perfect detection would create a system which generates nothing.  We cannot know the true size of $\mathcal{H^{+}}$ (see Figure~\ref{fig:ven_data}) without exhaustive manual hearing tests. We conduct blinded hearing tests of two generative systems. For \decfirst, a network which learns from spectrogram data utilizing CNN and LSTM layers is used. For \decsecond, we categorize the drum types with 3 different models: FC (Fully-Connected architecture and spectrograms for learning), CNN (CNN and LSTM layers with spectrograms), and E/F (Fully-Connected architecture with frequency and envelope features). The MEM system uses our best MEM, which simultaneously classifies sounds as drums and categorizes them.

@@

The percentage of sounds labeled as \enquote{bad} by the authors is a measurement of success with regard to \decfirst. We measure success with regard to \decsecond---the reliability of agreement between persons and drum categorization models---via the Fleiss' kappa coefficient \cite{fleiss1971measuring}. The value of 0 or less for this coefficient indicates no agreement beyond random chance, and the value of 1 indicates perfect agreement. We re-measure this coefficient after dropping all \enquote{bad} and \enquote{other} samples.

@@

\section{Dataset Details and Downloads}
\label{appendix:datasets}
\subsection{Download Instructions}
FreeDB, survey data, and randomly generated sounds can be downloaded from: \url{https://zenodo.org/record/3994999}\\\\
RadarDB requires lengthy downloads and processing. The script for its automated creation can be found under the \enquote{getting\_data} directory of the project:\\
\url{https://github.com/imilas/Synths_Stacks_Search} 
\\
Also included on the project page is our full source-code, models, and extra output examples.
\\\\
The latex source for this paper can be found at:\\
\url{https://github.com/imilas/CSMC-MuMe-2020}

\subsection{Details}
Our data sources are a large set of 2 second or shorter drum samples aggregated from personal libraries, free drum kits from the sample-swap project \footnote{https://sampleswap.org/} which we organized into 5 groups, and a large set of drum sounds aggregated from royalty free sources such as music radar \footnote{https://www.musicradar.com/}. We put together 3 databases of drums using these sources. We also created a database of synthetic noise from 1, 3, and 5 stacked virtual synthesizers. Specifications about our datasets can be found in table~\ref{table:all_db}.

@@

Using the optuna optimization tools \cite{akiba2019optuna}, we conducted 500 search trials. The trial's success is measured in their final loss value, calculated by applying the model to the test data-set. Each trial consisted of 20 epochs of training.
